Wind Veer and Speed in Turbulent Ekman Flow - Part 1: Scaling analysis and universal profile model

### 2024-10-31 write1/manuscript.tex
worked on comments from Hauke

The Coriolis force bends the path of motion on a rotating sphere and establishes geostrophic equilibrium when in balance with a pressure gradient force. Wind veer away from the wind direction in geostrophic equilibrium is
(i) due to direct frictional effects in the very vicinity of the surface and
(ii) due to turbulence which exerts indirect frictional effects; these effects
cause a slow-down of the mean wind reducing the Coriolis force thus turning the wind
in favor of the pressure gradient force. 
%%
Not only does the veering set the frame of reference for surface layer theory, it also has
effects at small and large scales from large-scale dispersion via plume spreading to
cyclone spin-down \citep{svensson:BM2009}
and on the capabilities of data assimilation and accuracy of surface flux estimates \citep{brown:QJR2005}. 
%
From a large-scale perspective, the veering of wind across the planetary boundary layer determines
the amount of cross-isobaric mass-flux, commonly referred to as 'Ekman pumping' \citep{ekman:AMA1905},
and it is thus a key factor in the life-cycle of large-scale synoptic systems.
%
% %% This shear is probably not related to Ekman-turning 
% On the meso-scale, shear in the surface layer a crucial factor for the initiation and sustained strong convective events (REFERENCE?).
% 
Within the atmospheric boundary layer (ABL), directional shear of the wind in
the upper part of the surface layer may cause a systematic yaw for tall wind power generation devices
where blades reach into the Ekman layer, i.e. that part of the boundary layer where the wind starts to turn;
an exact estimate of such effects is critical in the site assessments for wind farms
\citep{calaf:PF2010, mirocha:WES2018}. 
%
\par
%
In the planetary boundary layer, wind veer is characterized by the surface veering angle $\alpha$
defined as the angle between the negative surface shear stress $\tau_\SFC$ and the geostrophic wind.
%
Surface veering $\alpha$ and geostrophic drag $Z\equiv u_\star/G$, where the friction velocity $u_\star\equiv\sqrt{|\tau_\SFC|/\rho}$,
uniquely determine the surface drag $\tau_\SFC$ in a turbulent Ekman flow. 
%
In any quantitative description of the surface layer, the friction velocity $u_\star$ is the dynamic scale
and $\alpha$ defines the alignment of the frame of reference.
%
Knowledge about $u_\star$ and $\alpha$ is thus a prerequisite for any quantitative theory of the surface layer, 
and \citet{rossby:PIP1935} constrained the two parameters based on integral relations in the ABL.
%
Asymptotic similarity theory was later used by \cite{tennekes:JAS1973, blackadar:JAS1968},
and--based on his seminal direct numerical simulations (DNS) of Ekman flow--,~\cite{spalart:JFM1989} suggested
a modification to take into account effects of low to intermediate Reynolds numbers.
Later on, constants were re-evaluated with a focus on the ABL based on
observations \citep{hogstrom:BM1988,hogstrom:BM1996}
and numerical modelling \citep{spalart:PF2008, spalart:PF2009, ansorge:BM2014, ansorge:BM2019}.
%
\par
%
Attempts were also undertaken to obtain profiles of the wind speed: 
One approach is to match the inner and outer layer at a reference height;
\cite{etling:2002, emeis:2018}  (Sec. 21.10; Eq. 21.48) choose the Prandtl-layer height $z_\text{Prandtl}$ to match
the wind speed profiles, which, however, requires external prescription of $\alpha(z_\text{prandtl})$, the veering at that height. 
A one-dimensional profile with constant veering is given by \citet[Sec. 3; Eq. 3.1-3.19]{emeis:m2007}. 
%%%

@@

additional assumptions, such as a prescribed profile shape for eddy viscosity \cite{ellison:QJR1955} and two-layer models of the ABL take into account rotational effects at higher altitudes,

@@

a particular concern when it comes to wind-power forecasting \citep{optis:BM2014}.

@@

\section{Problem Formulation and Numerical Approach}

@@

\subsection{Notation and Governing Equations}

@@

The dimensional velocity vector of the numerical simulations is $\underline{U} = (U_1,U_2,U_3) = (U,V,W)$ over the coordinate system
$Oxyz$, where an approximate alignment (plus/ minus few degrees) of the direction $Ox$ with the surface shear stress is achieved.

@@

We denote the modulus of the surface shear, the surface friction, by

@@

Another common measure of scale separation is the Reynolds number defined by the laminar Ekman layer
thickness $D=\sqrt{2\nu/f}$, 
  \RE_D = \frac{G D} {\nu } = \sqrt{2 \RE_\Lambda}. 
Another common measure of scale separation is the Reynolds number defined by the laminar Ekman layer
thickness $D=\sqrt{2\nu/f}$, 
  \RE_D = \frac{G D} {\nu } = \sqrt{2 \RE_\Lambda}. 

@@

  \frac{\partial u_i}{\partial t} &=& \frac{\p \pi}{\p x_i} - u_j \frac{\p u_i}{\p x_j} + \epsilon_{i2j} (u_j -G_j)   + \frac{1}{\RE} \frac{\p^2 u_i}{\p x_{j}^2} \\ 

@@

\subsection{Numerical Simulations}

@@

(we find the expected slight decrease of $u_\star$ and $\alpha_\star$ with increasing $\RE$).

@@

      Re & $\delta^+=\RE_\tau$& $u_\star$ & $\alpha_\star$ & $\delta_{95}/\delta$ & $f u_\star^{-3}\int_{0}^{\delta} e \mathrm{d} z $ & $ G^{-3}\int_{0}^{\delta} \epsilon \mathrm{d}z$\\

@@

Velocity profiles in inner units ($U^+(z^+)$, Fig.~\ref{fig:profiles}a)and outer units ($U^-(z^-)$, Fig.~\ref{fig:profiles}b)

@@

      \textbf{a)} Shear-aligned velocity profiles in inner units;
      \textbf{b)} Shear-aligned velocity deficit in outer units; circles mark the
      height $z^-=0.15$; dashed line gives the best logarithmic fit to the data. 

@@

    inner \textbf{(c)} and outer height \textbf{(d)}.

@@

    as in Fig.~\ref{fig:profiles}.

@@

  S_\mathrm{turb} = \sqrt{\overline{uw}^2 + \overline{vw}^2}. 

@@

  \caption{horizontal slices of turbulence kinetic energy in the

@@

    outer layer (i=400);

@@

\end{figure} 
The organization of the flow is depicted in terms of the turbulence kinetic energy in Fig.~\ref{fig:slices}.
%
In vicinity of the wall, at $y^+\approx 10$, (Fig.~\ref{fig:slices}a), elongated streaks aligned with the
surface shear stress dominate.
%
Moving away from the wall, to $y^+\approx 150$ (well within the logarithmic region), the structures
are larger and more isotropic, but they are still largely aligned with the surface shear stress.
%
In the upper part of the outer layer, around $y^+\approx 1000$, no clear signature of the
surface veering direction is found, and intense TKE structures (bright yellow)
are organized on a large spatial scale with weaker eddies (greenish structures) and quiescent regions
in between.

@@

\subsection{Drag-Law}

@@

with  $z_\mathrm{ek} = \delta_\mathrm{ek} (y^- -s_\mathrm{ek})$.

@@

Thus, the reduction of the area under the Ekman hodograph is directly linked to the variation
in the friction velocity, which quantifies the well-known qualitative observation that a more turbulent boundary
layer leads to a flatter hodograph in Ekman flow.

@@

\begin{figure}
  \begin{flushleft}
    \textbf{\ \ a)}\hspace{0.45\textwidth}\textbf{\ \ b)}\\
  \end{flushleft} 
  \includegraphics[width=0.5\textwidth]{../plot/outer_layer_u.pdf}\includegraphics[width=0.5\textwidth]{../plot/outer_layer_w.pdf}
  \caption{
    \label{fig:profiles_outer}
    Shear-aligned velocity deficit for the wall-streamwise (panel \textbf{(a)}) and wall-spanwise (panel \textbf{(b)})
    components of the mean velocity $U^\alpha$ and $V^\alpha$.
    Solid lines show DNS data, dashed lines the Ekman profiles $U_\mathrm{ek}$ and $V_\mathrm{ek}$ as defined in
    Eq.~\ref{eqn:profile_ekman}. Region below $z^-=0.3$ is shown opaque. Variations in $U_\mathrm{ek}$ and $V_\mathrm{ek}$ are
    a consequence of the normalization and related to chagnes in $u_\star$ and $\alpha_\star$ among the different $\RE_D$. 
  }
\end{figure}

\subsection{Streamwise Velocity Component}
\label{sec:streamwise} 
For the streamwise velocity profile (that in non-rotating flows due to the geometry is always aligned with the surface shear stress),
well-established theories exist for various regimes according to their distance from the wall and the relative role of
viscosity, turbulence and interaction with the outer region of the flow with the logarithmic law for the mean velocity as a central anchor point.
%
\par
%
\begin{subequations}

@@

We therefore introduce a two-step correction procedure, accounting for the smaller-than linear growth beyond $y^+\approx 5$, and assuring smooth matching with the logarithmic law at $y^+=40$:
  U_\text{inner}^{{\alpha_\star}^+} = \frac{z^+}{1+c_1 (y^+)^2} +  (c_2 y^+ - a_\mathrm{match} )\frac{1+tanh\left[0.2(y^+-22)\right]}{2} + c_3 e^{-c_4(y^+-22)^2}. \label{eqn:u_inner} 

@@

The second and third terms on the right hand side vanish for $y^+\ll 22$, and $c_1=0.00185$
implies an approximately 5\% correction at $y^+=5$ and an 18.5\% correction at $y^+=10$.

@@

  {U_\text{log}^{\alpha_\star}}^+ = \frac{1}{\kappa} \log y^+ + C 

@@

For this logarithmic law, $a_\mathrm{match}=3.569861$ for a matching at $y^+=40$. 

@@

of $\alpha_\star$ and $u_\star$ implies that this difference cannot be constant as a function of $\RE$

@@

    \textbf{\ \ a}\hspace{0.45\textwidth}\textbf{\ \ b}\\

@@

  \caption{Profiles of shear-aligned span-wise velocity $(W^\alpha)^+$ versus inner height. 

@@

    Left panel show standard inner normalization. Right panel shows the inviscid normalization yielding

@@

@@

  \frac{\p}{\p z^+} \left[ f_{V,\mathrm{log}}\right]_{z^+=10} &=& \frac{\p}{\p z^+}\left[ F_{V,\mathrm{visc}}\right]_{z^+=10} &=:&d_{10}\simeq 4.01 \\

@@

\section{Discussion}
\subsection{Implications for surface-layer scaling} 
Eq.~(\ref{eqn:scaling_spanwise}) establishes a universal mixed scaling for the spanwise velocity
in the viscous layer:
While it requires the vertical coordinate to be expressed in inner units,
the velocity itself is normalized by the geostrophic wind, and becomes
inversely proportional to the friction Reynolds number $\RE_\tau=\delta^+$
when considered at a fixed height.
In vicinity of the surface, such mixed scaling has already been identified for higher-order statistics in convective 
flows \citep{mellado:BM2016,li:JAS2018}, where large scales leave their signatures in vicinity of the surface.
It is important to note here that, while $V$ is a first-order statistic from a statistical perspective,
the spanwise velocity is a higher-order correction term from the perspective of similarity theory and
from the viewpoint of the channel-flow analogy that is routinely employed in the surface layer.
Further, this is consistent with the scaling for the velocity hodograph found in
Eq.~(\ref{eqn:hodograph_scaling}) where the friction velocity also drops out.
In the surface layer, there is not only a mixed scaling--as we had already identified
in the viscous layer--, but we cannot find a universal function onto which the profiles
of spanwise velocity collapse. 
This additional degree of freedom reflects the inner--outer matching problem for the
spanwise velocity, and rather than giving a universal profile for this region,
as is usually done, we resort here to a parametric description of the problem,
namely in terms of the function $f_{V,\mathrm{log}}$ determined by the
parameters $a_\mathrm{log}$, $b_\mathrm{log}$, $c_\mathrm{log}$ which can be estimated
based on the above scaling considerations for any Reynolds number. 
We note that, once the parameter $a_\mathrm{log}$ is known, the parameters $b_\mathrm{log}$
and $c_\mathrm{log}$ can be estimated solely based on $f_{V,\mathrm{visc}}$, i.e. using the
value $v_{10}$ and $d_{10}$ found for the viscous region of the flow.
For the range of Reynolds number relevant to geophysical problems ($10^4\lesssim \RE_D \lesssim 10^6$),
the variation of $c_\mathrm{log}$ is, however, rather small.

\subsection{Comparison with other theories}
%%% VAN DRIEST SCALING
\begin{figure}
  \centerline{\includegraphics[width=0.66\textwidth]{../plot/vanDriest_profile.png}}
  \caption{Near-wall velocity profile according to the 
    van-Driest scaling (blue, solid) in comparison with 
    the present model (red, dashed), 
    the viscous law of the wall (orange dashed), and 
    the logarithmic law (green, solid)} 
  \label{fig:vanDriest}
\end{figure}  An alternative approach that considers viscous effects close to the surface is the van-Driest scaling \citep{vandriest:JOT1956}, 
where an exponential damping of Prandtl's mixing length is considered near the wall to yield 
\begin{align}
  \frac{\p u^+}{\p y^+} = \frac{2}{ 1+ \sqrt{1+\left(2\kappa y^+\right)^2 (1-\exp\left[-y^+\right])}};
\end{align}
the spanwise component is zero as no rotational effects are considered. 
%
Comparing our proposed formulation (Fig.~\ref{fig:vanDriest}) for the stream-wise velocity in the inner layer to  \citeauthor{vandriest:JOT1956}'s formulation 
yields later convergence of the velocity onto the logarithmic profile while, over all, it serves as an excellent model of the streamwise 
velocity component:
% 
Notable deviations (on the order of few percent) only occur in the region $10 < z^+ < 30$, 
where the velocity transitions from the linear to the logarithmic profile.

%EKMAN SPIRAL 
For the higher layers of the ABL, the Ekman spiral is the simplest model available. When employed across the entirety of the ABL, the spiral of a turbulent Ekman layer is flattened with respect to Ekman's lamniar solution, which  corresponds to a reduction of the veering angle both at the surface and throughout the ABL. We, however, find that a modified version of the Ekman spiral explicitly taking into account the surface boundary condition, is a consistent model and yields excellent agreement with the velocity profiles from DNS (Sec.~\ref{sec:ekman}).

@@

  \textbf{(a)\hspace{0.48\textwidth}(b)}\\
  \includegraphics[width=0.5\textwidth]{../plot/etling_direction_1.pdf} 
  \includegraphics[width=0.5\textwidth]{../plot/etling_magnitude_1.pdf}
  \caption{Comparison of the DNS data (dashed lines) with the two-layer model proposed by Etling (solid lines) for different Reynolds number. Left panel shows the relative wind veer, right panel shows the velocity magnitude. The etling model is calibrated by the surface veer from the DNS and roughness parameter for the Etling model is chosen according to the correction factor $\exp{\kappa A}$ such that the total veer and velocity magnitude agree. DNS profile in the buffer layer, defined here as $z^+<30$ are shown as opaque dashed lines as the two-layer model does not consider viscous effects.}

@@

\end{figure}A two-layer model consolidating both the logartihmic and Ekman layer can be obtained following the arguments by \cite{etling:2008}, cf. \cite{emeis:2018}. Given a surface veering and a matching height (extent of the logarithmic layer), a formulation for the velocity profile across both the logarithmic and the outer layer is obtained. A comparison using the surface veering based on our model and a matching height of $z_\text{prandtl}=0.05\delta$, which gives better results than the matching height of $0.1\delta$ suggested by \cite{etling:2008}, is shown in Fig.~\ref{fig:etling}. 

@@

While the K-Profiles concern the total stress, a consistent formulation of the turning would also require its partitioning to the individual components, i.e. 
the orientation of the stress vector $(\overline{u'w'}, \overline{v'w'})$ in the horizontal plane. If K-theory shall be used, this stress vector needs to be anti-parallel to the corresponding stress vector $(\partial_z U, \partial_z V)$. 
Fig.~\ref{fig:alignment} shows the direction of the velocity, gradient and stress 
vectors across the boundary layer. It turns out that the negative stress vector with respect to the wind direction and the flux vectors (absolute) have an approximately similar direction. 

@@

  \textbf{(a)\hspace{0.48\textwidth}(b)}\\

>>>>

The Coriolis force bends the path of motion on a rotating sphere and establishes geostrophic equilibrium when in balance with a pressure gradient force. Wind veer away from the wind direction in geostrophic equilibrium is
(i) due to direct frictional effects in the very vicinity of the surface and
(ii) due to turbulence which exerts indirect frictional effects; these effects
cause a slow-down of the mean wind reducing the Coriolis force thus turning the wind
in favor of the pressure gradient force. 
%%
Not only does the veering set the frame of reference for surface layer theory, it also has
effects at small and large scales from large-scale dispersion via plume spreading to
cyclone spin-down \citep{svensson:BM2009}
and on the capabilities of data assimilation and accuracy of surface flux estimates \citep{brown:QJR2005}. 
%
From a large-scale perspective, the veering of wind across the planetary boundary layer determines
the amount of cross-isobaric mass-flux, commonly referred to as 'Ekman pumping' \citep{ekman:AMA1905},
and it is thus a key factor in the life-cycle of large-scale synoptic systems.
%
% %% This shear is probably not related to Ekman-turning 
% On the meso-scale, shear in the surface layer a crucial factor for the initiation and sustained strong convective events (REFERENCE?).
% 
Within the atmospheric boundary layer (ABL), directional shear of the wind in
the upper part of the surface layer may cause a systematic yaw for tall wind power generation devices
where blades reach into the Ekman layer, i.e. that part of the boundary layer where the wind starts to turn;
an exact estimate of such effects is critical in the site assessments for wind farms
\citep{calaf:PF2010, mirocha:WES2018}. 
%
\par
%
In the planetary boundary layer, wind veer is characterized by the surface veering angle $\alpha$
defined as the angle between the negative surface shear stress $\tau_\SFC$ and the geostrophic wind.
%
Surface veering $\alpha$ and geostrophic drag $Z\equiv u_\star/G$, where the friction velocity $u_\star\equiv\sqrt{|\tau_\SFC|/\rho}$,
uniquely determine the surface drag $\tau_\SFC$ in a turbulent Ekman flow. 
%
In any quantitative description of the surface layer, the friction velocity $u_\star$ is the dynamic scale
and $\alpha$ defines the horizontal alignment of the frame of reference, i.e. the rotation of the surface 
friction with respect to the outer wind direction. 
%
Knowledge about $u_\star$ and $\alpha$ is thus a prerequisite for any quantitative theory of the surface layer, 
and \citet{rossby:PIP1935} constrained the two parameters based on integral relations in the ABL.
%
Asymptotic similarity theory was later used by \cite{tennekes:JAS1973, blackadar:JAS1968},
and--based on his seminal direct numerical simulations (DNS) of Ekman flow--,~\cite{spalart:JFM1989} suggested
a modification to take into account effects of low to intermediate Reynolds numbers.
Later on, constants were re-evaluated with a focus on the ABL based on
observations \citep{hogstrom:BM1988,hogstrom:BM1996}
and numerical modelling \citep{spalart:PF2008, spalart:PF2009, ansorge:BM2014, ansorge:BM2019}.
%
\par
%
Attempts were also undertaken to obtain profiles of the wind speed: 
One approach is to match the inner and outer layer at a reference height as usggested by
\cite{etling:2002, emeis:2018}  (Sec. 21.10; Eq. 21.48); they choose the Prandtl-layer height $z_\text{Prandtl}$ to match
the wind speed profiles, which, however, requires external prescription of $\alpha(z_\text{prandtl})$, the veering at that height. 
A one-dimensional profile with constant veering is given by \citet[Sec. 3; Eq. 3.1-3.19]{emeis:m2007}. 
%%%

@@

additional assumptions. One option is a prescribed profile shape for eddy viscosity (\cite{ellison:QJR1955}), another are two-layer models of the ABL that take into account rotational effects at higher altitudes,

@@

a particular concern when it comes to wind-power forecasting \citep{etling:2002, emeis:2018, optis:BM2014}.

@@

\section{Problem formulation and numerical approach}

@@

\subsection{Notation and governing equations}

@@

The dimensional velocity vector of the numerical simulations is $\underline{U} = (U_1,U_2,U_3) = (U,V,W)$ over the coordinate system $Oxyz$, where an approximate alignment (plus/ minus few degrees) of the direction $Ox$ with the surface shear stress is achieved.
We consider velocity profiles only, i.e. all velocities are averaged over horizontal planes and in time, that is, they 
correspond to an Ensemble average. 

@@

We denote the square root of the modulus of surface shear, the surface friction, by

@@

Another common measure of scale separation is the Reynolds number $\RE_D = G D /\nu $  defined by the laminar Ekman layer thickness $D=\sqrt{2\nu/f}$.

@@

  \frac{\partial u_i}{\partial t} &=& \frac{\p \pi}{\p x_i} - u_j \frac{\p u_i}{\p x_j} + \epsilon_{i2j} (u_j -g_j)   + \frac{1}{\RE_\Lambda} \frac{\p^2 u_i}{\p x_{j}^2} \\ 

@@

\subsection{Numerical simulations}

@@

(we find the expected slight decrease of $u_\star/G$ and $\alpha_\star$ with increasing $\RE$).

@@

      Re & $\delta^+=\RE_\tau$& $u_\star/G$ & $\alpha_\star$ & $\delta_{95}/\delta$ & $f u_\star^{-3}\int_{0}^{\delta} e \mathrm{d} z $ & $ G^{-3}\int_{0}^{\delta} \epsilon \mathrm{d}z$\\

@@

Velocity profiles in inner units ($U^{\alpha+}(z^+)$, Fig.~\ref{fig:profiles}a)and outer units ($U^{\alpha-}(z^-)$, Fig.~\ref{fig:profiles}b)

@@

      \textbf{a)} Shear-aligned velocity profiles in inner units; circles mark the
      height $z^-=0.15$;
      \textbf{b)} Shear-aligned velocity deficit in outer units; diamonds mark the grid point closest to the height $z^+=50$

@@

    inner height \textbf{(c)} and outer height \textbf{(d)}. \textbf{(c)} uses inner normalization; 
    \textbf{(d)} uses normalization by $\nu \delta^+$ which approximately collapses data in the 
    outer layer. 

@@

    as in Fig.~\ref{fig:profiles}

@@

  S_\mathrm{turb} = \sqrt{\overline{u' w'}^2 + \overline{v'w'}^2}, 
where dashed quantities $u'$, $v'$, $w'$ indicate deviations from the mean and the overbar denotes 
horizontal and time averaging. 

@@

  \caption{Horizontal slices of turbulence kinetic energy in the

@@

    outer layer (i=400) of the case with $Re_\tau=2978$;

@@

\end{figure} 
The organization of the flow with $Re_\tau=2978$ is depicted in terms of the turbulence kinetic energy in Fig.~\ref{fig:slices}.
%
In vicinity of the wall, at $y^+\approx 10$, (Fig.~\ref{fig:slices}a), elongated streaks aligned with the
surface shear stress dominate.
%
Moving away from the wall, to $y^+\approx 150$ (well within the logarithmic region), the structures
are larger and more isotropic, but they are still largely aligned with the surface shear stress.
%
In the upper part of the outer layer, around $y^+\approx 1000$, no clear signature of the
surface veering direction is found, and intense TKE structures (bright yellow)
are organized on a large spatial scale with weaker eddies (greenish structures) and quiescent regions
in between.

@@

\subsection{Drag-law}

@@

with  $z_\mathrm{ek} = \delta_\mathrm{ek} (z^- -s_\mathrm{ek})$.

@@

%Thus, the reduction of the area under the Ekman hodograph is directly linked to the variation
%in the friction velocity, which quantifies the well-known qualitative observation that a more turbulent boundary
%layer leads to a flatter hodograph in Ekman flow.

@@

\begin{figure}
  \begin{flushleft}
    \textbf{\ \ a)}\hspace{0.45\textwidth}\textbf{\ \ b)}\\
  \end{flushleft} 
  \includegraphics[width=0.5\textwidth]{../plot/outer_layer_u.pdf}\includegraphics[width=0.5\textwidth]{../plot/outer_layer_w.pdf}
  \caption{
    \label{fig:profiles_outer}
    Shear-aligned velocity deficit for the streamwise (panel \textbf{(a)}) and spanwise (panel \textbf{(b)})
    components of the mean velocity $U^\alpha$ and $V^\alpha$.
    Solid lines show DNS data, dashed lines the Ekman profiles $U_\mathrm{ek}$ and $V_\mathrm{ek}$ as defined in
    Eq.~\ref{eqn:profile_ekman}. 
    %Region below $z^-=0.3$ is shown opaque. 
    Variations in $U_\mathrm{ek}$ and $V_\mathrm{ek}$ are
    a consequence of the normalization and related to changes in $u_\star$ and $\alpha_\star$ among the different $\RE_D$. 
  }
\end{figure}

\subsection{Streamwise velocity component}
\label{sec:streamwise} 
Well-established theories exist for the streamwise velocity profile, which in non-rotating flows is aligned with the surface shear stress due to the geometry. These theories cover various regimes based on their distance from the wall and the relative influence of viscosity, turbulence, and interaction with the outer flow region, with the logarithmic law for the mean velocity serving as a central reference point.
%
\par
%
\begin{subequations}

@@

We therefore introduce a two-step correction procedure, accounting for the smaller-than linear growth beyond $z^+\approx 5$, and assuring smooth matching with the logarithmic law at $z^+=40$:
  U_\text{inner}^{{\alpha_\star}^+} = \frac{z^+}{1+c_1 (y^+)^2} +  (c_2 z^+ - a_\mathrm{match} )\frac{1+tanh\left[0.2(z^+-22)\right]}{2} + c_3 e^{-c_4(z^+-22)^2}. \label{eqn:u_inner} 

@@

The second and third terms on the right hand side vanish for $z^+\ll 22$, and $c_1=0.00185$
implies an approximately 5\% correction at $z^+=5$ and an 18.5\% correction at $z^+=10$.

@@

  {U_\text{log}^{\alpha_\star}}^+ = \frac{1}{\kappa} \log z^+ + C 

@@

For this logarithmic law, $a_\mathrm{match}=3.569861$ for a matching at $z^+=40$. 

@@

of $\alpha_\star$ and $u_\star$ implies that this difference cannot be constant as a function of $\RE$.
%\par 

@@

@@

  \caption{Profiles of shear-aligned span-wise velocity $(V^\alpha)^+$ versus inner height. 

@@

    Panel \textbf{(a)} shows standard inner normalization, panel \textbf{(b)} the inviscid normalization yielding

@@

  \label{eqn:match}

@@

  \frac{\p}{\p z^+} \left[ f_{V,\mathrm{log}}\right]_{z^+=10} &=& \frac{\p}{\p z^+}\left[ f_{V,\mathrm{visc}}\right]_{z^+=10} &=:&d_{10}\simeq 4.01 \\

@@

\subsection{Matching of the inner and outer layer profiles}
The formulations introduced above are continuous across the transition from the inner to 
the outer layer. 
However, the requirement of smooth derivatives would over-constrain the velocity profiles and is hence not applied. 
This can, in particular for low or extremely high $\RE$ cause discontinuity in the derivatives of the velocity profiles around the transition from the inner to the outer layer. 
To avoid such artificial discontinuity, the profiles are blended by an error-function transition using a blending 
height $z_{blend}^-=0.28-2.25\sqrt{1./Re_\tau}$ and a transition thickness 
$z_\text{trans}=2$, 
such that the weighting function $\omega(z^-)$ becomes 
\[
  \omega(z^-) = \frac12 \left[\text{erf}\left(z_\text{trans}\log\frac{z^-}{z_{\text{blend}}^{-}}\right) +1\right].
\]
hence 
 $u_\text{total} = (1-\omega) \times u_\text{inner} + \omega \times u_\text{outer}$ and similar for $v$.

%
\begin{figure}
  \phantom{AAA}\textbf{(a)\hspace{0.3\textwidth}(b)\hspace{0.3\textwidth}(c)}\\
  \includegraphics[width=\textwidth]{../plot/ekman_profiles.pdf}
  \caption{ \textbf{(a)} Velocity deficit, \textbf{(b)} velocity profile in shear-aligned hodographs and \textbf{(c)} hodograph in geostrophy-aligned coordinates.
   In panels \textbf{(a)} and \textbf{(b)}, blue lines correspond to the streamwise component and red to the spanwise. }
  \label{fig:summary} 
\end{figure} 
The resulting velocity profiles across the entire boundary layer are shown in inner and outer scaling as well as in ho\-do\-graph- view in Fig.~\ref{fig:summary}. The shear-aligned velocity deficit is shown in outer scaling highlighting the universality in the outer layer. 
% 
The logarithmic scaling of the streamwise component is encountered as straight blue lines in panels \textbf{(a)} and \textbf{(b)} where 
the extent of the logarithmic range increases with $\RE$ towards lower values of $z^-$ and higher values of $z^+$ depending on the scaling. 
Importantly, the logarithmic region is widening for increasing Reynolds number -- irespective of the scaling. 
% 
In this simple inner scaling, the spanwise velocity (which follows a mixed scaling) 
Does not collapse but seems to depend on $Re$ (the collapse is seen in Fig.~\ref{fig:inner_w}).
% 
However, the velocity deficit in outer units becomes approximately universal, also across the inner layer; 
this reflects the compensation of reduced turning ($\alpha$) by increased drag ($u_\star$), 
and is consistent with the theoretical considerations discussed in \ref{sec:drag-law}. 
% 
\par
%
The spanwise velocity at a fixed height scales approximately as $Re_\tau^{-1}$ (Sec.~\ref{sec:spanwise}).
%
However, the fraction  of turning that is encountered within the inner layer of the flow amounts to about 1/3 of the total wind veer  (Fig.~\ref{fig:etling}a).  
%
This is because, in inner units, the inner layer grows as $Re_\tau$  which exactly compensates the reduced gradient of spanwise velocity. 
% 
The hodographs show the well-known Ekman shape with the laminar profile as an outer limit and 'flatter'  hodographs, corresponding to less turning, for increasing Reynolds number.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Discussion}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Implications for surface-layer scaling} 
Eq.~(\ref{eqn:scaling_spanwise}) establishes a universal mixed scaling for the spanwise velocity in the viscous layer:
%
While it requires the vertical coordinate to be expressed in inner units, the velocity itself is normalized by the geostrophic wind, and becomes inversely proportional to the friction Reynolds number $\RE_\tau=\delta^+$ when considered at a fixed height.
%
In vicinity of the surface, such mixed scaling has already been identified for higher-order statistics in convective flows \citep{mellado:BM2016,li:JAS2018}, where large scales leave their signatures in vicinity of the surface.
%
It is important to note here that, while $V$ is a first-order statistic from a statistical perspective, the spanwise velocity is a higher-order correction term from the perspective of similarity theory and from the viewpoint of the channel-flow analogy that is routinely employed in the surface layer.
%
Further, this is consistent with the scaling for the velocity hodograph found in Eq.~(\ref{eqn:hodograph_scaling}) where the friction velocity also drops out.
%
\par
%
In the surface layer, there is not only a mixed scaling--as we had already identified in the viscous layer--, but we cannot find a universal function onto which the profiles of spanwise velocity collapse. 
%
This additional degree of freedom reflects the inner--outer matching problem for the spanwise velocity. 
%
Rather than giving a universal profile for this region, we resort here to a parametric description of the problem in terms of the function $f_{V,\mathrm{log}}$ determined by the parameters $a_\mathrm{log}$, $b_\mathrm{log}$, $c_\mathrm{log}$ which can be estimated based on the above scaling considerations for any Reynolds number.
%
We note that, once the parameter $a_\mathrm{log}$ is known, the parameters $b_\mathrm{log}$ and $c_\mathrm{log}$ can be estimated solely based on $f_{V,\mathrm{visc}}$, i.e. using the value $v_{10}$ and $d_{10}$ found for the viscous region of the flow.
%
For the range of Reynolds number relevant to geophysical problems ($10^4\lesssim \RE_D \lesssim 10^6$), the variation of $c_\mathrm{log}$ is, however, rather small.

\subsection{Comparison with other theories}
%%% VAN DRIEST SCALING
\begin{figure}
  \centerline{\includegraphics[width=0.66\textwidth]{../plot/vanDriest_profile.png}}
  \caption{Near-wall velocity profile according to the 
    van-Driest scaling (blue, solid) in comparison with 
    the present model (red, dashed), 
    the viscous law of the wall (orange dashed), and 
    the logarithmic law (green, solid)} 
  \label{fig:vanDriest}
\end{figure}  An alternative approach that considers viscous effects close to the surface is the van-Driest scaling \citep{vandriest:JOT1956}, 
where an exponential damping of Prandtl's mixing length is considered near the wall to yield 
\begin{align}
  \frac{\p u^+}{\p z^+} = \frac{2}{ 1+ \sqrt{1+\left(2\kappa z^+\right)^2 (1-\exp\left[-z^+\right])}};
\end{align}
the spanwise component is zero as no rotational effects are considered. 
%
Comparing our proposed formulation for the stream-wise velocity in the inner layer to  \citeauthor{vandriest:JOT1956}'s formulation yields later convergence of the velocity onto the logarithmic profile while, over all, it serves as an excellent model of the streamwise velocity cmponent (Fig.~\ref{fig:vanDriest}):
% 
Notable deviations (on the order of few percent) only occur in the region $10 < z^+ < 30$,  where the velocity transitions from the linear to the logarithmic profile. 
%
\par
%
%EKMAN SPIRAL 
For the higher layers of the ABL, the Ekman spiral is the simplest model available. 
%
When employed across the entirety of the ABL, the spiral of a turbulent Ekman layer is flattened with respect to Ekman's laminar solution, which  corresponds to a reduction of the veering angle both at the surface and throughout the ABL. 
%
We, however, find that a modified version of the Ekman spiral explicitly taking into account the surface boundary condition, is a consistent model and yields excellent agreement with the velocity profiles from DNS (Sec.~\ref{sec:ekman}).

@@

  \textbf{(a)\hspace{0.48\textwidth}(b)}\\[-0.15em]
  \includegraphics[width=0.5\textwidth]{../plot/etling_direction_0.png} 
  \includegraphics[width=0.5\textwidth]{../plot/etling_magnitude_0.png}
  \caption{Comparison of the DNS data (dashed lines) with the two-layer model proposed by Etling (solid lines) for different Reynolds number. Left panel shows the relative wind veer, right panel shows the velocity magnitude. The Etling model is calibrated by the surface veer from the DNS and the roughness parameter is chosen according to the correction factor $\exp{\kappa A}$ such that the total veer and velocity magnitude agree. Profiles in the buffer layer, defined here as $z^+<30$ are shown as opaque dashed lines as the two-layer model does not consider viscous effects.}

@@

A two-layer model consolidating both the logartihmic and Ekman layer can be obtained following the arguments by \cite{etling:2008}, cf. \cite{emeis:2018}. Given a surface veering and a matching height (extent of the logarithmic layer), a formulation for the velocity profile across both the logarithmic and the outer layer is obtained. A comparison using the surface veering based on our model and a matching height of $z_\text{prandtl}=0.05\delta$, which gives better results than the matching height of $0.1\delta$ suggested by \cite{etling:2008}, is shown in Fig.~\ref{fig:etling}. 

@@

While the K-Profiles concern the total stress, a consistent formulation of the turning would also require its partitioning to the individual components, i.e. the orientation of the stress vector $(\overline{u'w'}, \overline{v'w'})$ in the horizontal plane. If K-theory shall be used, this stress vector needs to be anti-parallel to the corresponding stress vector $(\partial_z U, \partial_z V)$. 
Fig.~\ref{fig:alignment} shows the direction of the velocity, gradient and stress  vectors across the boundary layer. It turns out that the negative stress vector with respect to the wind direction and the flux vectors (absolute) have an approximately similar direction. 

@@

  \textbf{(a)\hspace{0.48\textwidth}(b)}\\[-0.15em]
